---
title: "Environmental Spatial Statistics Final Project"
author: "Mohannad Zahra"
date: "2023-02-09"
output: html_document
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## Decreasing groundwater levels in Brandenburg:

## Abstract:

Generally, the change in groundwater level is caused by multiple factors, which makes separating single driving forces more difficult.
We focus on changes in the water budget in Berlin-Brandenburg, Germany's capital region, over the last five years. Data were compiled by
CORRECTIV. Measurements of groundwater levels in Brandenburg during the months of June, July, and August were used. This analysis was 
carried out to (1) identify spatial trends or covariates and their interactions with observed groundwater trends, (2) to identify local 
hydrological changes, and (3) to create a `model Variogram` and develop a `Kriging model` that helps us predict groundwater trends
spatially. Studies show that the Berlin-Brandenburg region is widely characterized by decreasing groundwater and lake levels.



## Introduction:

Climate change affects groundwater levels by increasing aridity and excessive water withdrawal from groundwater bodies. Regional
assessments of hydro-environmental changes have been carried out through spatial interpolation studies of measured changes in groundwater
levels by CORRECTIV in Berlin-Brandenburg during the last five years, and the results can generally be presented spatially through easily
interpretable digital elevation model maps and study and analyze the spatial groundwater trends.



## Study site:

In the northeast of Germany, the Brandenburg region is located. The state has an area of 29,480 square kilometres, making it the fifth
largest in Germany. Together with Berlin, it forms the Berlin-Brandenburg Metropolitan Region.
The Elbe and Oder are the two major rivers that drain into the North and Baltic seas, respectively. The Oder River flows in the eastern
part of Brandenburg, while the Elbe flows in the western part.
Rainfall in Brandenburg is relatively low, with a regional average of 604 mm per year, while evapotranspiration is relatively high, with a
regional average of 511 mm. *(1951-2000, Gerstengarbe al. 2003).*

## Motheds

## Load the library:

```{r}
library(spatstat)
library(gstat)
library(sp)
library(rgdal)
library(raster)
library(rgeos)
library(sf)
library(stars)
library(tmap)
```

## Read the data:

```{r}
DWD <-read.table('groundwater/groundwater_5year_trend.csv',sep=",",header = T)
head(DWD)
```

## We use unique function to  returns a data frame, but with duplicate elements/rows removed:

```{r}
UN<- unique(DWD$ID)
length(UN)
```

## Clean the data and remove the NaN values:

```{r}
DWWD <- DWD[complete.cases(DWD), ]
length(DWWD$ID)
```

## Reproject the data:

```{r}
coordinates(DWWD) <- ~ Lon + Lat
proj4string(DWWD) <- CRS("+init=epsg:4326")
DWDD <- spTransform(DWWD, CRS("+init=epsg:32632"))
summary(DWDD)
```
## Visualization the data:

```{r}
tmap_mode("view")
tm_shape(DWDD,name ="Groundwater Levels") + tm_dots(col = "Trend" ,size = 0.1,title ="Groundwater Levels") +
tm_tiles("Stamen.TonerLabels") + tm_basemap("Stamen.Watercolor")
```

## Convert our data frame  to an sf object:

```{r}
data <- st_as_sf(DWDD, coords = c("Lon", "Lat"))
```

## Distribution of the data:

```{r}
plot(data["Trend"], key.pos = 4, axes = TRUE, key.width = lcm(1.7), key.length = 1, main="Fig 2:  Groundwater Levels classification")
```


## Data feature:

According to our figure, we can see that most of our point distribution is between -0.2 and 0 and the mean value is around -0.05 so we can
know that most of these points have decreased during the time study.


## Point Pattern Analysis:

```{r}
water_ppp=as.ppp((data))
water_ppp =unmark(water_ppp)
```

## Testing the groundwater distribution:

### Estimating homogeneous intensity:

Quadrat counts tile the observation window into equal sized quadrats and then count the number of points falling into each tile.

```{r}
Q = quadratcount(water_ppp)

V <- distmap(water_ppp)
#plot(V,main="Fig 4: Distance", axes = TRUE)

par(mfrow=c(1,2)) 
plot(Q, main="Fig 3: Quadrat counting")
plot(data,add=T,col="blue",cex = .3)
plot(V,main="Fig 4: Distance", axes = TRUE)
plot(data,add=T,col="red",cex = .3)
```
### Kernel densities:

We can use this function to visually explore densities.
```{r}
D <- density(water_ppp, adjust = 0.1)
plot(D,main="Fig 5: spatial density", axes = TRUE)
```



According to figure 3 on the map above we can notice that the spatial density has a high concentration of water ground points in the east
and west central region of Brandenburg.



## Quadrat counting test of homogeneity:

The function quadrat.test tests this hypothesis.

```{r}
Groundwater_test <- quadrat.test(Q)
Groundwater_test
```


As *P_Value* the probability the variables are independent is less than the significance level (0.05), this means that our data is not a
homogeneous Poisson process `non-randomly spatially distributed` reflecting a trend in the groundwater deposits occurences. The approach
used is a dispersion test based on Quadrat counts, it performs a goodness-of-fit test of a fitted inhomogenous Poisson model.


## Relative distribution estimate:

```{r}
rh <- rhohat(water_ppp,D)
plot(rh, xlab = "Density between Groundwater Level", main="Fig 6: Estimate of the intensity ρ(D)")
```



Most of the water ground deposits has a distance between 3e-07 and 4e-07. This function estimates the kernel intensity , and it is
expressed as a function of the spatial covariate.



```{r}
rhv <- rhohat(water_ppp,log(V))
plot(rhv, xlab = "Distance to nearest Groundwater Level", main="Fig 7: Estimate of the intensity ρ(distance)")
```



The plot above is an estimate of the intensity ρ(X) as a function of distance. It indicates the waterground are relatively unlikely to be
found at great distance of the central region of the study area.

## We convert the Trend numbers from meter to millimeter to be easer to deal with:

```{r}
Trend.MM <- (DWWD$Trend)*1000
DWWD$Trend.MM = Trend.MM
head(DWWD)
```



## Plot Histogram:

```{r}
hist(log10(DWWD$Trend),xlab="Concentration",col = "blue",main="Fig 8: Groundwater levele Histogram")
```



We obtained a Gaussian distribution for the groundwater point and the largest intensity points around -1

## Calculate the experimental variogram and explore whether the variogram is isotropic or anisotropic.

### Variogram analysis:

```{r}
v <- variogram(Trend.MM~ 1,DWDD)
plot(v, col="blue", main="Fig 9: Groundwater levele Experimental variogram ")
```
```{r}
v <- variogram(Trend.MM ~ 1,DWDD,width= 1200)
plot(v ,main="Fig 10: The Experimental variogram with cutoff= 45000,width= 1200")
```
## Fitting the spherical variogram with nugget:

```{r}
vs <- fit.variogram(v, vgm(,"Sph",nugget=700))
plot(v, vs, main = "Fig 11: The Spherical Variogram of the groundwater Leveles with Nugget", col="red", lwd =2)
```
## Fitting the Gaussian variogram with nugget:

```{r}
vg <- fit.variogram(v, vgm(,"Gau",nugget=700))
plot(v, vg, main = "Fig 12: The Gaussian Variogram of the groundwater Leveles with Nugget", col="red",lwd =2)
```
## Fitting the Exponential variogram with nugget:

```{r}
ve <- fit.variogram(v, vgm(,"Exp",nugget=700))
plot(v, ve, main = "Fig 13: The Exponential Variogram of the groundwater Leveles", col="red", lwd =2)
```
## Sum of squared errors:

```{r}
# Spherical variogram 
sse_s <- attr(vs,"SSErr")
sse_s
```
```{r}
# Gaussian variogram 
sse_g <- attr(vg,"SSErr")
sse_g
```
```{r}
# Exponential variogram 
sse_e <- attr(ve,"SSErr")
sse_e
```

The result of sum of squared errors show that the Exponential variograms has the lowest sum of squared errors (38.70752).


## Akaike Information Criterion (AIC):

AIC is an estimator of out-of-sample prediction error but includes more parameters in the role of estimating differences between observed
and modeled data, so AIC is scientifically more accurate than SSE.

```{r}
n <- length(v)

# Spherical
AIC_s <- length(v)*log(sse_s/n) +2*1
AIC_s
```
```{r}
# Gaussian
AIC_g <- length(v)*log(sse_g/n) +2*1
AIC_g
```
```{r}
# Exponential 
AIC_e <- length(v)*log(sse_e/n) +2*1
AIC_e
```

From the estimator AIC results we can see that the Exponential variogram has the lowest AIC value [13.18565].

## read the DEM for Brandenburg area:
```{r}
DEMM = read_stars("output_COP90.tif")
DEM_e = DEMM$output_COP90.tif
```



## DEM analysis:

### Quadrats determined by covariate:

We used tesselate space using quantiles of the digital elevation model to test whether a elevation (from DEM) has an impact on the spatial
distribution of Groundwater levels.

```{r}
dem = as.im(DEM_e)
b <- quantile(dem,probs=(0:5)/5, type=2)
elevcut <- cut(dem,breaks=b, labels=1:5)
Vt <- tess(image=elevcut)

plot(Vt,main="Fig 14: Elevation testing")
```
### Quadrat test:

```{r}
Qdem <- quadratcount(water_ppp,tess=Vt)
tSdem <- quadrat.test(Qdem)

tSdem
```
```{r}
par(mfrow=c(1,2))
plot(Qdem, main="Fig 15: Quadrant DEM")
plot(tSdem, main="Fig 16: Quadrant test for DEM and groundwater", alpha=0.5)
```


## Ordinary Kriging:

```{r}
# Interpolation 
k = krige(Trend.MM ~ 1 , location= DWWD,newdata = DEMM, model = ve,nmax = 5)
```

```{r}
summary(k)
```

## These results could be visualized as a map:

```{r}
library(ggplot2)

# Prediction map
k%>% as.data.frame %>%
ggplot(aes(x=x, y=y)) + geom_tile(aes(fill=var1.pred)) + coord_equal() +
         scale_fill_gradient(low = "yellow", high = "red") +
         ggtitle("Fig 17: Prediction of Groundwater (mm/year)")+
         guides(fill=guide_legend(title='Prediction (mm/year)'))+
         theme_bw()
```

```{r}
# variance map
k%>% as.data.frame %>%
ggplot(aes(x=x, y=y)) + geom_tile(aes(fill=var1.var)) + coord_equal() +
         scale_fill_gradient(low = "yellow", high = "red") +
         ggtitle("Fig 18: Variance of Groundwater (mm/year)")+
         guides(fill=guide_legend(title='Variance'))+
         theme_bw()
```




## Conclusion:

Based on the kriging map (Figure 17), it was possible to verify the distribution of the water ground content in the regions with higher and
lower concentrations in the evaluated area.


According to (Stock, M. & W. Lahmer (2007)) this study illustrates that the reason behind this decrease in Groundwater levels because of
the decline in precipitation up to the middle of this century and, the mean daily temperatures which increased not only in summer but also
in winter.





## References:

1. Stock, M. & W. Lahmer (2007): Brandenburg: A region in Germany suffering from water deficit - In spite of its numerous lakes. In: Lozán,
J. L., H. Grassl, P. Hupfer, L. Menzel & C.-D. Schönwiese.


2. Sonja Germer, Knut Kaiser, Oliver Bensand Reinhard F. Hüttl (2011) Water Balance Changes and Responses of Ecosystems and Societyin the
Berlin-Brandenburg Region – a Review.
